Comparison of FDR methods on 16S microbiome data with SummarizedBenchmark

### 2018-01-19 datasets/microbiome/microbiome-SummarizedBenchmark.Rmd
Update microbiome results.
Filter OTUs by presence in 1% of samples. Do genus-level analysis
for Baxter and Goodrich.

## Remove OTUs with fewere than 10 reads
#otu <- remove_shallow_otus(otu, 10)

@@

##### Phylogeny?!

@@

```{r, schubert-phylo, fig.width=10, fig.height=6, eval=TRUE}
#strat_hist(res, pvalue="pval", covariate="mean_abun_present", maxy=20)
#rank_scatter(res, pvalue="pval", covariate="ubiquity")
ggplot(res, aes(x=pval)) + geom_histogram() + facet_wrap(~phylum, scales = "free")
```

@@

@@

```{r, goodrich-phylo, fig.width=10, fig.height=6}
#strat_hist(res, pvalue="pval", covariate="mean_abun_present", maxy=20)
#rank_scatter(res, pvalue="pval", covariate="ubiquity")
ggplot(res, aes(x=pval)) + geom_histogram() + facet_wrap(~phylum, scales = "free")
```

@@

@@

@@

```{r, baxter-cleandata}
remove_shallow_smpls <- function(df, n_reads) {
  # Removes samples with fewer than n_reads from dataframe df.
  # df has OTUs in rows and samples in columns
  return(df[, colSums(df) > n_reads])
  }

@@

## Remove OTUs with fewere than 10 reads
#otu <- remove_shallow_otus(otu, 10)

@@

@@

```{r, baxter-phylo, fig.width=10, fig.height=6}
#strat_hist(res, pvalue="pval", covariate="mean_abun_present", maxy=20)
#rank_scatter(res, pvalue="pval", covariate="ubiquity")
ggplot(res, aes(x=pval)) + geom_histogram() + facet_wrap(~phylum, scales = "free")
```

@@

>>>>

## Remove OTUs with fewer than 10 reads
otu <- remove_shallow_otus(otu, 10)

@@

##### Phylogeny

@@

```{r, schubert-phylo, fig.width=10, fig.height=6, eval=FALSE}
#strat_hist(res, pvalue="pval", covariate="mean_abun_present", maxy=20)
#rank_scatter(res, pvalue="pval", covariate="ubiquity")
ggplot(res, aes(x=pval)) + geom_histogram() + facet_wrap(~phylum, scales = "free")
```

@@

remove_rare_otus <- function(df, perc_samples){
  return(df[rowSums(df > 0) / dim(df)[2] > perc_samples, ])
}

## Remove OTUs with fewer than 10 reads
otu <- remove_shallow_otus(otu, 10)
print(dim(otu))
## Remove OTUs in fewer than 1% of samples
otu <- remove_rare_otus(otu, 0.01)
print(dim(otu))

# Remove samples with fewer than sample_reads reads
sample_reads <- 100
otu <- remove_shallow_smpls(otu, sample_reads)

@@

```{r, goodrich-phylo, fig.width=10, fig.height=6, eval=FALSE}
#strat_hist(res, pvalue="pval", covariate="mean_abun_present", maxy=20)
#rank_scatter(res, pvalue="pval", covariate="ubiquity")
ggplot(res, aes(x=pval)) + geom_histogram() + facet_wrap(~phylum, scales = "free")
```

@@

## Goodrich, genus-level analysis

```{r, goodrich-genus}
## Collapse baxter OTU table to genus level

# Add column with otu names
genus_df <- as.data.frame(abun_otu)
genus_df$otu <- rownames(genus_df)
# Melt into tidy format
genus_df <- genus_df %>% melt(id.var="otu", variable.name="sample", value.name="abun")
# Split into separate taxonomic levels
genus_df <- genus_df %>% 
  separate(otu, c("kingdom", "phylum", "class", "order", "family", "genus", "species", "denovo"), 
           sep=";", remove = FALSE)

# Get rid of unannoated genera, and sum abundances for genera
genus_df <- genus_df %>% 
  filter(genus != "g__") %>% 
  group_by(genus, sample) %>% 
  summarise(total_abun = sum(abun))

# Convert back to longform
genus_df <- genus_df %>% spread(genus, total_abun) %>% as.data.frame
rownames(genus_df) <- genus_df$sample
genus_df <- genus_df %>% select(-sample) %>% t

# And re-order columns to match metadata
genus_df <- genus_df[, match(meta$X, colnames(genus_df))]
```

```{r, goodrich-genus-pvals}
resfile <- paste0(resdir, "goodrich_results_",
                  nrow(genus_df), "_genera.RData")
if (!file.exists(resfile)){
  # Get case and control indices
  case_idx <- meta$DiseaseState %in% c("OB")
  ctrl_idx <- meta$DiseaseState %in% c("H")

  casedf <- genus_df[, case_idx]
  ctrldf <- genus_df[, ctrl_idx]
  for (o in rownames(genus_df)) {
    # wilcoxon p value  
    w <- wilcox.test(casedf[o, ], 
                     ctrldf[o, ])
    p <- w$p.value
    teststat <- w$statistic

    # standard error of the OTU abundance, across all samples
    ses <- c(ses, 
             sd(genus_df[o, ])/sqrt(length(genus_df[o, ]))
             )

    # mean OTU abundance across all samples (after removing pseudo-count)
    mean_abuns <- c(mean_abuns, 
                    mean(genus_df[o, ] - zeroabun)
                    )

    # mean OTU abundance across only samples with the OTU present
    mean_abuns_present <- c(mean_abuns_present, 
                            sum(genus_df[o, ] - zeroabun) / sum(genus_df[o, ] > zeroabun)
                            )

    # ubiquity of OTU across all samples
    ubis <- c(ubis, 
              sum(genus_df[o, ] > zeroabun) / length(genus_df[o, ])
              )

    # effect (logfold difference)
    effs <- c(effs, 
              log(mean(genus_df[o, ctrl_idx])/mean(genus_df[o, case_idx]))
              )

  res <- data.frame(otu = rownames(genus_df), 
                    pval = pvals, test_statistic = teststats,
                    SE = ses, effect_size = effs, 
                    mean_abun = mean_abuns, mean_abun_present = mean_abuns_present,
                    ubiquity = ubis)
  res <- res %>%
    mutate(test_statistic = qnorm(exp(log(pval)-log(2)), lower.tail=FALSE) * sign(effect_size))

}else{
  load(resfile)
}
```

```{r, goodrich-genus-ubi, fig.width=10, fig.height=3.5}
strat_hist(res, pvalue="pval", covariate="ubiquity", maxy=10, binwidth=0.05)
rank_scatter(res, pvalue="pval", covariate="ubiquity")
```

```{r, goodrich-genus-abun, fig.width=10, fig.height=3.5}
strat_hist(res, pvalue="pval", covariate="mean_abun_present", maxy=8, binwidth=0.05)
rank_scatter(res, pvalue="pval", covariate="mean_abun_present")
```

```{r, goodrich-genus-labelcovariate}
res <- res %>% mutate(ind_covariate = ubiquity)
```

```{r, goodrich-genus-rm-ash}
res <- res %>% 
  mutate(test_statistic = NA) %>%
  mutate(effect_size = NA)
```

```{r, goodrich-genus-sb, results="hide", message=FALSE, eval=TRUE}
for (nmids in c(5, 50, 150)){
  print(nmids)
  bd <- initializeBenchDesign(nmids=nmids)

  resfile <- paste0(resdir, "goodrich_summarizedBenchmark_",
                    nrow(res), "_genera_nmids_", nmids, ".RData")
  duration <- NA
  if (!file.exists(resfile)){
    t1 <- proc.time()
    sb <- bd %>% buildBench(data=res, ftCols = "ind_covariate")
                            #parallel=TRUE, BPPARAM=multicoreParam)
    metadata(sb)$data_download_link <- "https://zenodo.org/record/840333/files/crc_baxter_results.tar.gz"
    save(sb, file=resfile)
    duration <- round((proc.time()-t1)[3]/60,1)
  }else{
    load(resfile)
  }
}

```{r, goodrich-genus-metrics, eval=TRUE}
# rename assay to qvalue
assayNames(sb) <- "qvalue"
sb <- addDefaultMetrics(sb)
```

```{r, goodrich-genus-plot, width=15, height=15, eval=TRUE}
# plot nrejects by method overall and stratified by covariate
rejections_scatter( sb, as_fraction=FALSE, supplementary=FALSE)
rejection_scatter_bins(sb, covariate="ind_covariate", supplementary=FALSE)
#plotFDRMethodsOverlap(sb, alpha=0.1, supplementary=FALSE, order.by="freq", nsets=100 )
```

Note: Benjamini-Hochberg (bh) overlaps exactly with the IHW results.

```{r, goodrich-genus-overlap}
plotFDRMethodsOverlap(sb, alpha=0.1, supplementary=FALSE, order.by="freq", nsets=100 )
```

@@

A brief aside: what's the count distribution of these OTUs?

```{r, eval=FALSE}
otu %>% colSums %>% hist()
```

@@

```{r, baxter-cleandata}
remove_shallow_smpls <- function(df, n_reads) {
  # Removes samples with fewer than n_reads from dataframe df.
  # df has OTUs in rows and samples in columns
  return(df[, colSums(df) >= n_reads])
  }

@@

remove_rare_otus <- function(df, perc_samples){
  return(df[rowSums(df > 0) / dim(df)[2] > perc_samples, ])
}

## Remove OTUs with fewer than 10 reads
otu <- remove_shallow_otus(otu, 10)
print(dim(otu))
## Remove OTUs in fewer than 1% of samples
otu <- remove_rare_otus(otu, 0.01)
print(dim(otu))

# Remove samples with fewer than sample_Reads reads
sample_reads <- 100
otu <- remove_shallow_smpls(otu, sample_reads)
print(dim(otu))

@@

# Remove empty OTUs
otu <- otu[rowSums(otu) > 0, ]
print(dim(otu))

@@

```{r, baxter-phylo, fig.width=10, fig.height=6, eval=FALSE}
#strat_hist(res, pvalue="pval", covariate="mean_abun_present", maxy=20)
#rank_scatter(res, pvalue="pval", covariate="ubiquity")
ggplot(res, aes(x=pval)) + geom_histogram() + facet_wrap(~phylum, scales = "free")
```

@@

## Baxter, genus-level analysis

```{r, baxter-genus}
## Collapse baxter OTU table to genus level

# Add column with otu names
genus_df <- as.data.frame(abun_otu)
genus_df$otu <- rownames(genus_df)
# Melt into tidy format
genus_df <- genus_df %>% melt(id.var="otu", variable.name="sample", value.name="abun")
# Split into separate taxonomic levels
genus_df <- genus_df %>% 
  separate(otu, c("kingdom", "phylum", "class", "order", "family", "genus", "species", "denovo"), 
           sep=";", remove = FALSE)

# Get rid of unannoated genera, and sum abundances for genera
genus_df <- genus_df %>% 
  filter(genus != "g__") %>% 
  group_by(genus, sample) %>% 
  summarise(total_abun = sum(abun))

# Convert back to longform
genus_df <- genus_df %>% spread(genus, total_abun) %>% as.data.frame
rownames(genus_df) <- genus_df$sample
genus_df <- genus_df %>% select(-sample) %>% t

# And re-order columns to match metadata
genus_df <- genus_df[, match(meta$Sample_Name_s, colnames(genus_df))]
```

```{r, baxter-genus-pvals}
resfile <- paste0(resdir, "baxter_results_",
                  nrow(genus_df), "_genera.RData")
if (!file.exists(resfile)){
  # Get case and control indices
  case_idx <- meta$DiseaseState %in% c("CRC")
  ctrl_idx <- meta$DiseaseState %in% c("H")

  casedf <- genus_df[, case_idx]
  ctrldf <- genus_df[, ctrl_idx]
  for (o in rownames(genus_df)) {
    # wilcoxon p value  
    w <- wilcox.test(casedf[o, ], 
                     ctrldf[o, ])
    #w <- kruskal.test(list(casedf[o, ], ctrldf[o, ]))
    p <- w$p.value
    teststat <- w$statistic

    # standard error of the OTU abundance, across all samples
    ses <- c(ses, 
             sd(genus_df[o, ])/sqrt(length(genus_df[o, ]))
             )

    # mean OTU abundance across all samples (after removing pseudo-count)
    mean_abuns <- c(mean_abuns, 
                    mean(genus_df[o, ] - zeroabun)
                    )

    # mean OTU abundance across only samples with the OTU present
    mean_abuns_present <- c(mean_abuns_present, 
                            sum(genus_df[o, ] - zeroabun) / sum(genus_df[o, ] > zeroabun)
                            )

    # ubiquity of OTU across all samples
    ubis <- c(ubis, 
              sum(genus_df[o, ] > zeroabun) / length(genus_df[o, ])
              )

    # effect (logfold difference)
    effs <- c(effs, 
              log(mean(genus_df[o, ctrl_idx])/mean(genus_df[o, case_idx]))
              )

  res <- data.frame(otu = rownames(genus_df), 
                    pval = pvals, test_statistic = teststats,
                    SE = ses, effect_size = effs, 
                    mean_abun = mean_abuns, mean_abun_present = mean_abuns_present,
                    ubiquity = ubis)
  res <- res %>%
    mutate(test_statistic = qnorm(exp(log(pval)-log(2)), lower.tail=FALSE) * sign(effect_size))

}else{
  load(resfile)
}
```

```{r, baxter-genus-ubi, fig.width=10, fig.height=3.5}
strat_hist(res, pvalue="pval", covariate="ubiquity", maxy=10, binwidth=0.05)
rank_scatter(res, pvalue="pval", covariate="ubiquity")
```

```{r, baxter-genus-abun, fig.width=10, fig.height=3.5}
strat_hist(res, pvalue="pval", covariate="mean_abun_present", maxy=8, binwidth=0.05)
rank_scatter(res, pvalue="pval", covariate="mean_abun_present")
```

```{r, baxter-genus-labelcovariate}
res <- res %>% mutate(ind_covariate = ubiquity)
```

```{r, baxter-genus-rm-ash}
res <- res %>% 
  mutate(test_statistic = NA) %>%
  mutate(effect_size = NA)
```

```{r, baxter-genus-sb, results="hide", message=FALSE, eval=TRUE}
for (nmids in c(5, 50, 150)){
  print(nmids)
  bd <- initializeBenchDesign(nmids=nmids)

  resfile <- paste0(resdir, "baxter_summarizedBenchmark_",
                    nrow(res), "_genera_nmids_", nmids, ".RData")
  duration <- NA
  if (!file.exists(resfile)){
    t1 <- proc.time()
    sb <- bd %>% buildBench(data=res, ftCols = "ind_covariate")
                            #parallel=TRUE, BPPARAM=multicoreParam)
    metadata(sb)$data_download_link <- "https://zenodo.org/record/840333/files/crc_baxter_results.tar.gz"
    save(sb, file=resfile)
    duration <- round((proc.time()-t1)[3]/60,1)
  }else{
    load(resfile)
  }
}

```{r, baxter-genus-metrics, eval=TRUE}
# rename assay to qvalue
assayNames(sb) <- "qvalue"
sb <- addDefaultMetrics(sb)
```

```{r, baxter-genus-plot, width=15, height=15, eval=TRUE}
# plot nrejects by method overall and stratified by covariate
rejections_scatter( sb, as_fraction=FALSE, supplementary=FALSE)
rejection_scatter_bins(sb, covariate="ind_covariate", supplementary=FALSE)
#plotFDRMethodsOverlap(sb, alpha=0.1, supplementary=FALSE, order.by="freq", nsets=100 )
```

Note: Benjamini-Hochberg (bh) overlaps exactly with the IHW results.

```{r, baxter-genus-overlap}
plotFDRMethodsOverlap(sb, alpha=0.1, supplementary=FALSE, order.by="freq", nsets=100 )
```
